Kids First or Marriage First?

### 2019-05-22 main/draft.tex
Introduction version 1

\section{Introduction}
As of 2018, [40\%] of new births occurred by unmarried mothers. This number is rising for decades, but, more importantly, it is rather heterogeneous, reaching three-quarters for mothers who are African American, [...] under 25 and [...] for high school or less in education. But even for educated and older women this share is substantial [...]. An increase in non-marital cohabitation is also a documented trend, despite of this, persistent non-marital cohabitation is still not the most common thing: according to SIPP 2014, only [...] of women 21--40 cohabitate for longer than 5 years without being married. A substantial part of children by unmarried parents are born in such cohabitating unions (SIPP [...]).

These numbers have important economic implications: children born by unmarried women, as documented by, for instance [...], display significantly worse future outcomes in terms of education and income, and this pattern survives controlling in many dimensions. Although it is well understood that children of never married single mothers or children of blended families (when a woman marries someone else than child's father) have generally worse outcomes, even when parents marry each other after they made a baby their marriage is statistically less stable, as I argue in this paper, and the larger chances of marriage dissolution do affect future prospects of the children.

This pairs well with common perception that some couples marry at or shortly after the moment they make a baby: long-lasting marital cohabitations with children are even less common (among ...). If cohabitation is viewed as uncommitted relationship, unexpected fertility may serve as a trigger to obtain more commitment. Narrowly, we can interpret this as shotgun marriage story, however in the times with large abortion access many unplanned pregnancies are aborted, and cohabitating couples seem to \emph{choose} to keep an unplanned baby. The whole role of marriage changes as it is more socially acceptable to cohabitate, though it is much more conventional to do it before having kids. As highlighted by, for instance, [...], despite of general decrease of role of marriage in the society, securing investment to children is one of the most important forces that makes people marry each other.

@@

Namely, I document that females who had their first birth before their first marriage have higher chances of divorce in the future, this is mainly driven by women married to the parents of their children, and this is robust and relatively more pronounced for high education or high income groups. I then provide a model incorporating the bargaining mechanism that can rationalize these differences. I estimate the model parameters to match observable patterns in the data and show what it implies about the role of unequal exposure to childcare costs in observable marriage selection and divorce patterns. Shortly, the model suggests if people could perfectly control timing of their births, general share of divorced people would decrease by ..., especially in high-income groups.

Section 2 of this paper discuss existing results and general trends that are already described by the literature. Section [...] discuss how I construct the dataset and timing variables, and Section 3 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences. After that I explain theoretical model in Section 4, the process of setting its parameters in Section 5. Section 6 provides estimation results and counterfactual simulations implied by these results. Section 7 provides more general discussion and concludes.

%Understanding how people make marriage decisions and more generally how household structure is formed is crucial for analyzing labor market outcomes, assets accumulation, various dimensions of inequality and children development outcome. Most of existing studies, even ones that take household structure seriously, typically abstract from the fact that household structure is itself a choice variable. However, existing trends in marriage and divorce rates, increasing share of kids with unmarried parents and in blended families, and, especially, how different these trends are by income and racial groups --- this all suggests economic importance of household issues.

%In particular, economists know pretty little about how people decide on marriage and divorce, despite of the fact that these issues have lots of economic content. Stylized models describing marriage (and, in smaller extent, divorce) emerged in family economics, however most of them describe qualitative issues and do not deliver quantitative predictions that can be applied for analyzing policy or just understanding how different factors contribute to the recent changes. ...]

\section{Existing Results}

@@

The main conceptual framework that I utilize are models of household's dynamic bargaining and limited commitment, extensively reviewed by

@@

\newpage
\section{Robustness Checks}
\section{Conclusions}

\appendix
\section{Additional Empirical Results}
\subsection{Alternative Income Partitions\label{inc-part}}

>>>>

\section{Introduction}
As of 2018, [40\%] of new births occurred by unmarried mothers. This number is rising for decades, but, more importantly, it is rather heterogeneous, reaching three-quarters for mothers who are African American, [...] under 25 and [...] for high school or less in education. But even for educated and older women this share is substantial [...]. An increase in non-marital cohabitation is also a documented trend, despite of this, persistent non-marital cohabitation is still not the most common thing: according to National Survey of Family And Growth, median duration of cohabitation is 22 months, and median duration of cohabitation if it ends by marriage is 21.\footnote{Source: Copen et al, \cite{copen}}. Unmarried unions with children are still pretty rare: according to Payne, 2013\nocite{payne}, only about 3\% of children live in unions with unmarried cohabitating parents (as opposed to 21\% with single mothers).

These numbers have important economic implications: children born by unmarried women, as documented by, for instance Kearney, Levine, 2017, display significantly worse future outcomes in terms of education and income, and this pattern survives controlling in many dimensions. Although it is well understood that children of never married single mothers or children of blended families (when a woman marries someone else than child's father) have generally worse outcomes, even when parents marry each other after they made a baby their marriage is statistically less stable, as I argue in this paper, and the larger chances of marriage dissolution do affect future prospects of the children.

This pairs well with common perception that some couples marry at or shortly after the moment they make a baby. If cohabitation is viewed as uncommitted relationship, unexpected fertility may serve as a trigger to obtain more commitment. Narrowly, we can interpret this as shotgun marriage story, however in the times with large abortion access many unplanned pregnancies are aborted, and cohabitating couples seem to \emph{choose} to keep an unplanned baby. The whole role of marriage changes as it is more socially acceptable to cohabitate, though it is much more conventional to do it before having kids. As highlighted by, for instance, Lafortune, 2018, despite of general decrease of role of marriage in the society, securing investment to children is one of the most important forces that makes people marry each other.

@@

Namely, I document that females who had their first birth before their first marriage have higher chances of divorce in the future, this is mainly driven by women married to the parents of their children, and this is robust and relatively more pronounced for high education or high income groups. I then provide a model incorporating the bargaining mechanism that can rationalize these differences. I estimate the model parameters to match observable patterns in the data and show what it implies about the role of unequal exposure to childcare costs in observable marriage selection and divorce patterns. Shortly, the model suggests if people could perfectly control timing of their births, general share of divorced people would decrease by [...], especially in high-income groups.

Section 2 of this paper discuss existing results and general trends that are already described by the literature. Section 3 discuss how I construct the dataset and timing variables, and Section 5 establishes main observations about differences in groups depending on their marriage timing, together with a number of additional considerations useful to understand the content of these differences. After that I explain theoretical model in Section 5 and the process of setting its parameters in Section 6. Section 7 describes estimation results and counterfactual simulations implied by these results. Section 8 provides more general discussion and concludes.

\section{Related Literature}

@@

The main conceptual framework that I utilize are models of household's dynamic bargaining and limited commitment, extensively reviewed by Chiaporri, Mazzocco 2015.\nocite{chiappori-review} These models were used to answer various marriage and divorce related questions. Theoretical part of my model is mostly based on Voena, 2015\nocite{voena-1}, that uses dynamic bargaining to consider impact of changes in divorce laws, and Low et al, 2018\nocite{low-1}, focusing on marriage and divorce selection related to welfare utilitzation. There are few more recent papers utilitizing similar approaches, for instance, Shephard, 2019\nocite{shephard} studying gender and age pay gaps through equilibrium on marriage markets.

To model fertility choices and shocks, I rely on recent lifecycle models. I make many choices similar to Sommer, 2015\nocite{sommer} who studies how changes in reproductive horizon affect couple's fertility decisions with in the environment with income shocks, paying detailed attention to time and money structure of childcare. More detailed treatment of unintended pregnancies is done by Ejrnæs, Jørgensen, 2018\nocite{ejrnaes}, they extend the model to allow for non-random abortion decisions though do not model childcare costs in details.

Commitment in marriage and its interaction with fertility decisions attract recent attention as possible drivers of trends in marriage and divorce. Work by Lafortune, Low 2018\nocite{lafortune-1} uses ACS data and housing prices, as well as marriage market model, to argue that married people buy houses as a device to commit not to divorce, that allows them to secure investment in their future children. This is parallel and complimentary with my results, that mainly focus on consequences of not having enough commitment and how they relate to fertility timing.

Impact of non-marital childbearing and parents divorce are well-documented empirically from children prospective. For example, Kearney, Levine 2017 use PSID to document how being born out of marriage statistically child's outcome in school years and in early adult life, and the heterogeneity of this by parent's education and age groups. Shortly, they find presence of ``marriage premium'' for children at each education level, though for less educated and younger mothers benefits of having married parents are more pronounced in early life, and for elder and more educated mothers --- in adult life. A classic study by Akerlof, 1996\nocite{akerlof} highlight the mechanism why unplanned pregnancies may not be aborted in couples so they are still present even with large abortion access. Finally, Ginther, Pollak, 2004 show that outcomes of children in blended families (i.e. having step-parents) are similar to outcomes of children of single parents.

This paper fills an important gap in the literature considering impact of out-of-marriage births for women. As fertility, marriage and divorce involve many individual decisions affected by many unobservable factors, it is not possible to ask casual questions by using only the data, and existing empirical studies only provide descriptive evidence. This study contributes to the growing literature of quantitative models that both use the tools of family economics to endogenize timing of fertility, marriage, divorce and their interactions, and, on the other hand, provide numbers and patterns consistent with microdata, so they can be used to answer policy questions.

As noted by Ginther and Pollak, causal questions in the context of marriage and divorce should be taken with caution: to answer, for instance, how parents' divorce affect child's future we have to be precise what counterfactual we compare it with (either parents never divorced, parents stayed together for few more years, parents never married). The same issues concern impact of unplanned pregnancies in cohabitating couples, therefore, instead of trying to ask causal quesitons I the main goal of using the model is to deliver counterfactual outcomes, that still facilitate understanding of some causality.

@@

\section{Robustness Checks}
\section{Conclusions}
\clearpage
\section*{Literature}
\bibliography{bib}

\newpage
\appendix
\section{Additional Empirical Results}
\subsection{Alternative Income Partitions\label{inc-part}}
